---
title: "Data Visualization in R"
output: 
  html_document:
    theme: readable
---

Hana &#352;ev&#269;&#237;kov&#225;, Christy Lam

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
ntd.file <- "ntd-monthly-February2017.xls"
ev <- FALSE
```

# Setup

Install required packages by
```{r eval=FALSE}
install.packages(c("ggplot2", "plotly", "readxl"))
```

On your file system create a working directory and then navigate into it, e.g.:

```{r eval=FALSE}
getwd()
setwd("../rviz-workshop")
```

# Loading Dataset

We will work with a [dataset from the National Transit Database (NTD)](https://www.transit.dot.gov/ntd/data-product/monthly-module-adjusted-data-release). Define a function that can load such dataset for any month and year:

```{r eval=FALSE}
download.ntd <- function(month, year) {
    url <- "https://www.transit.dot.gov/sites/fta.dot.gov/files/"
    url.file <- paste0(url, month, "%20", year, "%20Adjusted%20Database.xls")
    local.file <- paste0("ntd-monthly-", month, year, '.xls')
    download.file(url.file, local.file, mode="wb")
    return(local.file)
}
```

The argument ``mode="wb"`` specifies that it is a binary file and thus, should be loaded in binary mode. If your file is an ASCII (text) file, remove that argument.

Load an NTD dataset for February 2017:

```{r eval=ev}
month <- "February"
year <- 2017
```

```{r eval=FALSE}
ntd.file <- download.ntd(month, year)
```

Read the UPT sheet:

```{r eval=ev, results="hide"}
library(readxl)
excel_sheets(ntd.file)
ds <- read_excel(ntd.file, sheet="UPT")
```

```{r eval=FALSE}
View(ds)
```

# Data Manipulation

Convert to data frame:

```{r eval=ev}
class(ds)
df <- as.data.frame(ds)
class(df)
dim(df)
```

Reduce the number of rows by extracting relevant operators:

```{r eval=ev}
operators <- c(1, 3, 5, 20, 23, 28, 29, 35, 40, 54)
df <- df[df[,1] %in% operators, ]
dim(df)
```

```{r eval=FALSE}
View(df)
```


Reduce the number of columns by removing the unnecessary ones:

```{r eval=ev}
df <- df[, -c(1:2, 4:7, 9)]
```

Use abbrevations for agency names:

```{r eval=ev}
abbr <- c("Central Puget Sound Regional Transit Authority"="ST",
          "King County Department of Transportation - Metro Transit Division"="KCM",
          "Everett Transit"="ET",
          "Kitsap Transit"="KT",
          "Pierce County Transportation Benefit Area Authority"="PT",
          "Snohomish County Public Transportation Benefit Area Corporation"="CT",
          "Pierce County Ferry Operations"="PCF",
          "City of Seattle - Seattle Center Monorail Transit"="SDOT",
          "Washington State Ferries"="WSF",
          "King County Ferry District"="KCF",
          "King County Department of Transportation"="KCF"
          )
df$Agency <- abbr[df$Agency]
```

Replace the Modes column by aggregated modes and sum values over them. Sort the resulting dataset by agency:

```{r eval=ev}
modes.lookup <- data.frame(Modes=c("CB","TB","MB", "LR","SR", "CR", 
                                   "FB", "VP", "MO","MG", "DR","DT"),
                   AModes=c("bus", "bus", "bus", "light_rail", "light_rail", 
                           "commuter_rail", "ferry", "vanpool", "monorail", "monorail", 
                           "demand_reponse", "demand_reponse"))
modes.lookup
df <- merge(df, modes.lookup, by="Modes")
df <- df[,-1]
df <- aggregate(. ~ AModes+Agency, df, sum)
df <- df[order(df$Agency),]
dim(df)
```

# Basic plots

Time series:

```{r eval=ev}
busdf <- subset(df, AModes=="bus")
rownames(busdf) <- busdf$Agency
busdf <- as.matrix(busdf[,-(1:2)])
n <- ncol(busdf)
plot(1:n, busdf["CT",], xlab="time", ylab="", main="UPT bus", 
     ylim=range(busdf), log="y", type='l') # try type="p" and type="b"
abline(v=seq(12.5, n, by=12), lty=2)
text(seq(1, n, by=36), min(busdf), seq(2002, 2017, by=3), pos=4)
lines(1:n, busdf["ET",], col="blue")
lines(1:n, busdf["KCM",], col="green")
```

Box plots, histograms, scatterplots:

```{r eval=ev}
boxplot(t(busdf))
hist(busdf["KT",])
plot(busdf["KT",], busdf["ET",])
abline(0,1)
```

# Preparing data for advanced graphics

```{r eval=ev}
library(reshape2)
months <- toupper(month.abb)
result <- NULL
# iterate over all years
for(year in 2002:2017) {
  # columns to extract for one year
  yearchar <- sprintf("%02d", year-2000)
  cols <- c("AModes", "Agency", paste0(months, yearchar))
  # extract only columns present in the dataframe
  subdf <- df[, cols[cols %in% colnames(df)]]
  # remove year from the month names
  colnames(subdf) <- gsub(yearchar, "", colnames(subdf))
  # convert to long format and add year column
  subdfl <- melt(subdf, id=c("AModes", "Agency"))
  subdfl$Year <- year
  # add to the resulting dataframe
  result <- rbind(result, subdfl)
}
```


# Use of Functions

Put the data manipulation tasks from above into a function:

```{r eval=ev}
process.ntd.dataset <- function(metrics="UPT") {
  abbr <- c("Central Puget Sound Regional Transit Authority"="ST",
          "King County Department of Transportation - Metro Transit Division"="KCM",
          "Everett Transit"="ET",
          "Kitsap Transit"="KT",
          "Pierce County Transportation Benefit Area Authority"="PT",
          "Snohomish County Public Transportation Benefit Area Corporation"="CT",
          "Pierce County Ferry Operations"="PCF",
          "City of Seattle - Seattle Center Monorail Transit"="SDOT",
          "Washington State Ferries"="WSF",
          "King County Ferry District"="KCF",
          "King County Department of Transportation"="KCF"
          )
  operators <- c(1, 3, 5, 20, 23, 28, 29, 35, 40, 54)
  modes.lookup <- data.frame(Modes=c("CB","TB","MB", "LR","SR", "CR", 
                                     "FB", "VP", "MO","MG", "DR","DT"),
                            AModes=c("bus", "bus", "bus", "light_rail", "light_rail", 
                                    "commuter_rail", "ferry", "vanpool", "monorail", "monorail", 
                                    "demand_reponse", "demand_reponse"))
    
  ds <- read_excel(ntd.file, sheet=metrics)
  df <- as.data.frame(ds)
  
  df <- df[df[,1] %in% operators, ]
  df <- df[, -c(1:2, 4:7, 9)]
  
  df$Agency <- abbr[df$Agency]

  df <- merge(modes.lookup, df, by="Modes")
  df <- aggregate(. ~ AModes+Agency, df, sum)
  return(df)
}

get.ntd.long <- function(df) {
  months <- toupper(month.abb)
  result <- NULL
  # iterate over all years
  for(year in 2002:2017) {
    # columns to extract for one year
    yearchar <- sprintf("%02d", year-2000)
    cols <- c("AModes", "Agency", paste0(months, yearchar))
    # extract only columns present in the dataframe
    subdf <- df[, cols[cols %in% colnames(df)]]
    # remove year from the month names
    colnames(subdf) <- gsub(yearchar, "", colnames(subdf))
    # convert to long format and add year column
    subdfl <- melt(subdf, id=c("AModes", "Agency"))
    subdfl$Year <- year
    # add to the resulting dataframe
    result <- rbind(result, subdfl)
  }
  return(result)
}
```

To create a data frame for the sheet "VRM", do

```{r eval=ev}
df <- process.ntd.dataset("VRM")
dfl <- get.ntd.long(df)
tail(dfl)
```
