---
title: "Data Visualization in R"
output: 
  html_document:
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
ntd.file <- "ntd-monthly-January2017.xls"
ev <- FALSE
```

# Setup

Install required packages by
```{r eval=FALSE}
install.packages(c("ggplot2", "plotly", "readxl"))
```

On your file system create a working directory and then navigate into it, e.g.:

```{r eval=FALSE}
getwd()
setwd("../rviz-workshop")
```

# Loading Dataset

We will work with a [dataset from the National Transit Database (NTD)](https://www.transit.dot.gov/ntd/data-product/monthly-module-adjusted-data-release). Define a function that can load such dataset for any month and year:

```{r eval=FALSE}
download.ntd <- function(month, year) {
    url <- "https://www.transit.dot.gov/sites/fta.dot.gov/files/"
    url.file <- paste0(url, month, "%20", year, "%20Adjusted%20Database.xls")
    local.file <- paste0("ntd-monthly-", month, year, '.xls')
    download.file(url.file, local.file)
    return(local.file)
}
```

Load an NTD dataset for January 2017:

```{r eval=ev}
month <- "January"
year <- 2017
```

```{r eval=FALSE}
ntd.file <- download.ntd(month, year)
```

Read the UPT sheet:

```{r eval=ev, results="hide"}
library(readxl)
excel_sheets(ntd.file)
df <- read_excel(ntd.file, sheet="UPT")
View(df)
```

TODO: Rename some columns and rows and show different types of visualization.
