---
title: "Leaflet in R"
output: 
  html_document:
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This exercise will take concepts from previous workshops such as manipulating demographic tabular data in excel and combine it with new concepts such as reading in spatial data and producing a choropleth map via [Leaflet for R](https://rstudio.github.io/leaflet/).

# Setup

* Install required packages if they have not already been installed: openxlsx, rgdal, sp, leaflet, RColorBrewer, htmlwidgets
```{r eval=FALSE}
install.packages(c("openxlsx", "rgdal", "sp", "leaflet", "RColorBrewer", "htmlwidgets"))
```

* Load libraries
```{r eval=FALSE}
library(openxlsx)
library(rgdal)
library(sp)
library(leaflet)
library(RColorBrewer)
library(htmlwidgets)
```

# Data Manipulation

* Using the ``download.file`` function, download data from the Office of Financial Management (OFM) at the block group level: http://www.ofm.wa.gov/pop/smallarea/data/xlsx/saep_bg10.xlsx

* Read-in the first sheet in the workbook containing Total Population estimates. Note where the row of column names begin. Experiment how changing the argument ``check.names`` to ``TRUE`` changes the column names.

* Subset the data to contain only records from the Puget Sound Region.

* Pare-down table so only select fields are available: 'County Name', 'Census Block Group Code Complete', 'Estimated Total Population 2000' through 'Estimated Total Population 2016'

* Rename fields to something shorter.

# Reading Shapefiles

* Read-in the block group shapefile in the 'W:' drive using the ``readOGR`` function from the rgdal package.

```{r eval=FALSE}
input.shp.dir <- 'W:/geodata/census/Blockgroup'
input.shp.name <- 'blockgrp2010'
bg10.shp <- readOGR(dsn = input.shp.dir, layer = input.shp.name)
```

* Check if successful:
    + To get a quick and dirty visual of your shapefile, type in the console: ``plot(bg10.shp)``
    + To view the attribute table of your spatial object, type the object name followed by ``@data``


* Leaflet is particular about projection. Re-project the block group shapefile to the Geographic Coordinate System 'WGS84' using the ``spTransform`` function from the sp package. You will need to use this Coordinate Reference System (CRS) statement as one of its arguments:
    + ``CRS("+proj=longlat +datum=WGS84")``

```{r eval=FALSE}
bg10.shp.proj <- spTransform(bg10.shp, CRS("+proj=longlat +datum=WGS84"))
```

* Join the tabular data to your projected block group object.

# Visualize with Leaflet for R
You'll find helpful hints on https://rstudio.github.io/leaflet/. The 'Basemap', 'Choropleths' and 'Legends' tabs will help, and you can always type ``?`` followed by a function name to access help via the console.

* Initialize leaflet with your data. You can include a basemap using the ``addProviderTiles`` function. One has been provided in the example below, but you can select a different one from the full set [here](http://leaflet-extras.github.io/leaflet-providers/preview/index.html). 
```{r eval=FALSE}
m <- leaflet(data = ) %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap) %>% 
  addPolygons()

print(m)
```

* To color block groups by Total Population for a given year, first define breakpoints to classify your data values. Try using the histogram function ``hist()`` to examine the range and distribution of your estimates.

* Store the breakpoints (including min and max values) in a vector and assign it to object ``bins``. This line should come before your leaflet object.

* To map your bins to colors, use the ``colorBin`` function.
    + The first argument of the ``colorBin`` function requires a color palette. You can use pre-defined color palettes from the package ``RColorBrewer``. Type ``display.brewer.all()`` or ``brewer.pal.info`` into the console to view your palette options.

* Set the ``colorBin`` function to object ``pal``. This line will also come before your leaflet object.

```{r eval=FALSE}
bins <- 
pal <- colorBin(palette =  , bins = bins)

```

* Modify ``addPolygons()`` to include style properties. 
    + The ``fillColor`` argument will reference your 'pal' object. The column name containing your data values will go inside the parenthesis.
    + The ``fillOpacity`` argument controls the levels of transparency for ``fillColor``. It accepts a numeric value between 0 and 1.
    + Experiment with other style properties such as ``stroke`` and ``weight``.
```{r eval=FALSE}

m <- leaflet(data = ) %>%
  addProviderTiles(providers$Esri.NatGeoWorldMap) %>% 
  addPolygons(
    fillColor = ~pal(),
    fillOpacity = 1
  )

print(m)
```

* You can add layers to your leaflet object by chaining another ``addPolygons()``. See if you can add another layer showing the top 5 block groups with the largest Total Population in 2016. Style their borders with a different color and remove the fill.
    + create a new spatial object containing only the top 5 block groups.
    + chain another ``addPolygons()`` and complete the following style property arguments
    + the ``color`` argument can be set to a [hex code](http://www.colorhexa.com/)
```{r eval=FALSE}
  addPolygons(data = ,
              fill = ,
              stroke = ,
              color = ,
              opacity = 
  )

```  

* Add a legend

* Additional items that can be added to your map:
    + a layers control box using the 'addLayersControl' function which allows you to turn layers on and off
    + pop-up information so that whenever a block group is clicked-on, information such as the population estimate will appear.

* Export your leaflet object as an html file. See the notes from the last [R workshop on Data Visualization](https://rawgit.com/psrc/R-tutorials/master/rviz-workshop.html)

*Extra Credit: in lieu of the layer showing the 5 block groups with the largest Total Population, display 5 block groups with the largest increases in population between 2015-16 and 5 block groups with the largest decreases*

### Additional Resources
* [Intro to visualising spatial data in R](http://spatial.ly/wp-content/uploads/2013/12/intro-spatial-rl-3.pdf)
* [Overview of Coordinate Reference Systems in R](https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/OverviewCoordinateReferenceSystems.pdf)
* [EPSG codes to project shapefiles](http://www.epsg-registry.org/)
